---
title: "Farmand et al 2026 - NeuroFit RCT Clinical and Cognitive Outcomes Analysis Examples"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
library(readxl)
library(dplyr)
library(tidyr)
library(glmmTMB)
library(nlme)
library(ordinal)
library(DHARMa)
library(emmeans)
library(ggplot2)
```

## 1. Import dataset

```{r import-data}
df <- read_excel("Farmand et al - NeuroFit_RCT_Clinical_Cognitive_Dataframe_ITT_Population.xlsx")
```

## 2. Reshape data to long format

```{r reshape-data}
ldat <- df %>%
  pivot_longer(
    cols = starts_with(c("A_", "B_")),
    names_to = c("time", ".value"),
    names_sep = "_"
  ) %>%
  mutate(
    Participant_ID = as.factor(Participant_ID),
    Intervention_group = factor(
      Intervention_group,
      levels = c(0, 1),
      labels = c("Control", "Exercise")
    ),
    time = factor(time, levels = c("A", "B")),
    Sex = as.factor(Sex),
    Education = as.factor(Education)
  )
```

## 3. Create baseline covariates

```{r baseline-covariates}
ldat <- ldat %>%
  group_by(Participant_ID) %>%
  mutate(
    baselineLDI = LDI[time == "A"],
    baselineREC = REC[time == "A"],
    baselineMCS12 = MCS12[time == "A"],
    baselinePCS12 = PCS12[time == "A"],
    baselinePHQ9 = PHQ9[time == "A"],
    baselineMOCA = MOCA[time == "A"],
    baselinePSQI = PSQI[time == "A"],
    baselineBMI = BMI[time == "A"],
    baselineBF = BF[time == "A"],
    baselineHR = HR[time == "A"],
    baselineFit = Fit[time == "A"],
    baselineSysBP = SysBP[time == "A"],
    baselineDiaBP = DiaBP[time == "A"],
    baselineWeight = Weight[time == "A"],
    baselineWtH = WtH[time == "A"]
  ) %>%
  ungroup()
```

## 4. Transform variables for beta regression

Beta regression was used for bounded continuous outcomes. Variables were min-max normalised to lie between 0 and 1, with a small epsilon adjustment to avoid exact 0 or 1 values.

Beta regression outcomes:

```{r beta-outcomes}
beta_outcomes <- c("LDI", "REC", "MCS12", "PCS12", "PHQ9", "BF")
```

```{r beta-transformation}
epsilon <- 1e-5

normalize_beta <- function(x) {
  min_val <- min(x, na.rm = TRUE)
  max_val <- max(x, na.rm = TRUE)

  scaled <- (x - min_val) / (max_val - min_val)
  adjusted <- scaled * (1 - 2 * epsilon) + epsilon

  return(adjusted)
}

ldat <- ldat %>%
  mutate(across(
    all_of(beta_outcomes),
    normalize_beta,
    .names = "n{col}"
  ))
```

## 5. Example beta GLMM

The following model illustrates the beta GLMM approach. The same model structure was applied to other beta-regression outcomes by replacing the dependent variable and corresponding baseline covariate.

```{r beta-glmm-example}
m_LDI <- glmmTMB(
  nLDI ~ Intervention_group * time + Sex + Age + Education + baselineLDI +
    (1 | Participant_ID),
  data = ldat,
  family = beta_family(link = "logit"),
  REML = FALSE
)

summary(m_LDI)
confint(m_LDI, parm = "beta_", level = 0.95, method = "Wald")
```

## 6. Beta GLMM diagnostics

```{r beta-diagnostics}
sim_LDI <- simulateResiduals(m_LDI)

plot(sim_LDI)
testDispersion(sim_LDI)
testUniformity(sim_LDI)
testOutliers(sim_LDI)
```

## 7. Example linear mixed-effects model

Linear mixed-effects models were used for normally distributed continuous outcomes.

LME outcomes:

```{r lme-outcomes}
lme_outcomes <- c(
  "MOCA", "BMI", "HR", "SysBP", "DiaBP",
  "Fit", "PSQI", "Weight", "WtH"
)
```

Example model using MOCA:

```{r lme-example}
m_MOCA <- lme(
  MOCA ~ Intervention_group * time + Sex + Age + baselineMOCA + Education,
  random = ~ 1 | Participant_ID,
  data = ldat,
  na.action = na.exclude
)

summary(m_MOCA)
intervals(m_MOCA)
```

## 8. LME diagnostics

```{r lme-diagnostics}
res_MOCA <- resid(m_MOCA, type = "normalized")
fit_MOCA <- fitted(m_MOCA)

plot(fit_MOCA, res_MOCA,
     xlab = "Fitted values",
     ylab = "Normalised residuals")
abline(h = 0, lty = 2)

qqnorm(res_MOCA)
qqline(res_MOCA)

hist(res_MOCA, main = "Residuals histogram", xlab = "Normalised residuals")
```

## 9. Ordinal mixed model for IPAQ

IPAQ was analysed using a cumulative link mixed model because it was an ordered categorical variable.

```{r ipaq-model}
ldat <- ldat %>%
  mutate(
    IPAQ = factor(IPAQ, levels = c(0, 1, 2), ordered = TRUE),
    time = factor(time, levels = c("A", "B"))
  )

m_IPAQ <- clmm(
  IPAQ ~ Intervention_group * time + Sex + Age + Education +
    (1 | Participant_ID),
  data = ldat,
  link = "logit",
  Hess = TRUE
)

summary(m_IPAQ)
```

## 10. Estimated marginal means

Estimated marginal means were calculated using `emmeans`. The following example uses the LDI beta GLMM.

```{r emmeans-example}
LDI_emm <- emmeans(m_LDI, ~ Intervention_group * time)

LDI_emm_summary <- summary(
  LDI_emm,
  infer = TRUE,
  type = "response"
)

LDI_emm_summary
```

## 11. Back-transform beta-regression estimates

For beta-regression outcomes, estimated marginal means were back-transformed to the original outcome scale.

```{r back-transform-beta}
min_LDI <- min(ldat$LDI, na.rm = TRUE)
max_LDI <- max(ldat$LDI, na.rm = TRUE)

back_transform_LDI <- function(z) {
  ((z - epsilon) / (1 - 2 * epsilon)) * (max_LDI - min_LDI) + min_LDI
}

LDI_emm_raw <- as.data.frame(LDI_emm_summary) %>%
  mutate(
    Mean = back_transform_LDI(response),
    Lower95 = back_transform_LDI(asymp.LCL),
    Upper95 = back_transform_LDI(asymp.UCL),
    Mean_CI = sprintf("%.2f [%.2f, %.2f]", Mean, Lower95, Upper95),
    Time = recode(time, A = "Baseline", B = "Endpoint")
  ) %>%
  select(
    Group = Intervention_group,
    Time,
    Mean,
    Lower95,
    Upper95,
    Mean_CI
  )

LDI_emm_raw
```

## 12. Example estimated marginal means plot

```{r emm-plot}
plot_df <- as.data.frame(LDI_emm_summary) %>%
  mutate(
    fit_orig = back_transform_LDI(response),
    ymin_orig = back_transform_LDI(asymp.LCL),
    ymax_orig = back_transform_LDI(asymp.UCL),
    time = factor(time, levels = c("A", "B")),
    Intervention_group = factor(
      Intervention_group,
      levels = c("Control", "Exercise")
    )
  )

ggplot(plot_df, aes(x = time, y = fit_orig, group = Intervention_group)) +
  geom_point(aes(colour = Intervention_group), size = 2) +
  geom_line(aes(colour = Intervention_group), linewidth = 0.25) +
  geom_ribbon(
    aes(ymin = ymin_orig, ymax = ymax_orig, fill = Intervention_group),
    alpha = 0.2
  ) +
  scale_x_discrete(labels = c("A" = "Baseline", "B" = "Endpoint")) +
  labs(
    x = "Time point",
    y = "LDI",
    colour = "Intervention group",
    fill = "Intervention group"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.title = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```