---
title: "Farmand et al 2026 - NeuroFit RCT Blood-Based Biomarker Analysis Examples"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
library(readxl)
library(dplyr)
library(tidyr)
library(nlme)
library(emmeans)
library(ggplot2)
```

## 1. Import dataset

```{r import-data}
df <- read_excel("Farmand et al - NeuroFit_RCT_Blood_Based_BioMarkers_Dataframe_ITT_Population.xlsx")
```

## 2. Reshape data to long format

```{r reshape-data}
ldat <- df %>%
  pivot_longer(
    cols = starts_with(c("A_", "B_")),
    names_to = c("time", ".value"),
    names_sep = "_"
  ) %>%
  mutate(
    Participant_ID = as.factor(Participant_ID),
    Intervention_group = factor(
      Intervention_group,
      levels = c(0, 1),
      labels = c("Control", "Exercise")
    ),
    time = factor(time, levels = c("A", "B")),
    Sex = as.factor(Sex),
    Education = as.factor(Education)
  )
```

## 3. Create baseline covariates

```{r baseline-covariates}
ldat <- ldat %>%
  group_by(Participant_ID) %>%
  mutate(
    baselineKlo = Klo[time == "A"],
    baselineCLU = CLU[time == "A"],
    baselineBDNF = BDNF[time == "A"]
  ) %>%
  ungroup()
```

## 4. Blood-based biomarker outcomes

Linear mixed-effects models were used for the blood-based biomarker outcomes.

```{r biomarker-outcomes}
biomarker_outcomes <- c("Klo", "CLU", "BDNF")
```

## 5. Example linear mixed-effects model

The following model illustrates the linear mixed-effects model used for blood-based biomarker outcomes. The same model structure was applied to the other biomarker outcomes by replacing the dependent variable and corresponding baseline covariate.

```{r lme-example}
m_Klo <- lme(
  Klo ~ Intervention_group * time + Sex + Age + baselineKlo + Education,
  random = ~ 1 | Participant_ID,
  data = ldat,
  na.action = na.exclude
)

summary(m_Klo)
intervals(m_Klo)
```

## 6. Model diagnostics

```{r lme-diagnostics}
res_Klo <- resid(m_Klo, type = "normalized")
fit_Klo <- fitted(m_Klo)

plot(
  fit_Klo,
  res_Klo,
  xlab = "Fitted values",
  ylab = "Normalised residuals"
)
abline(h = 0, lty = 2)

qqnorm(res_Klo)
qqline(res_Klo)

hist(
  res_Klo,
  main = "Residuals histogram",
  xlab = "Normalised residuals"
)
```

## 7. Estimated marginal means

```{r emmeans}
Klo_emm <- emmeans(m_Klo, ~ Intervention_group * time)

Klo_emm_summary <- summary(
  Klo_emm,
  infer = TRUE
)

Klo_emm_summary
```

## 8. Estimated marginal means plot

```{r emm-plot}
plot_df <- as.data.frame(Klo_emm_summary) %>%
  mutate(
    time = factor(time, levels = c("A", "B")),
    Intervention_group = factor(
      Intervention_group,
      levels = c("Control", "Exercise")
    )
  )

ggplot(plot_df, aes(x = time, y = emmean, group = Intervention_group)) +
  geom_point(aes(colour = Intervention_group), size = 2) +
  geom_line(aes(colour = Intervention_group), linewidth = 0.25) +
  geom_ribbon(
    aes(ymin = lower.CL, ymax = upper.CL, fill = Intervention_group),
    alpha = 0.2
  ) +
  scale_x_discrete(labels = c("A" = "Baseline", "B" = "Endpoint")) +
  labs(
    x = "Time point",
    y = "Klotho",
    colour = "Intervention group",
    fill = "Intervention group"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.title = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```
